Notes

### 2022-09-20 notes/notes.tex
simplify equation

\subsection{Fitting the model}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise natural logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\frac{1}{2}\norm{\log.(P_{det} M_{shift} W(t) \tilde{x}_{365})}^2 + \frac{1}{2}\lambda \norm{\log W}^2
.$$
The first term measures the average logarithm of the predicted values. The second term measures the average logarithm of the weights.
\begin{proposition}
    Let $z$ be the vector obtained by applying the function $x \mapsto (\log x) / x$ to all predicted concentrations $\hat{y}_i$ corresponding to detection events. Then the gradient of $f(W)$ is the diagonal matrix
    $$
    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left((P_{det} M_{shift})^T z \tilde{x}_{365}^T\right) + \lambda W^{-1} \log W
    $$
    where $\mathrm{Diag}$ is the map that sets the off-diagonal elements of a matrix to zero.
\end{proposition}
\begin{proof}[Proof (still to be checked for calculation errors)]

Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
Define $$
G(t) := 
P_{det} M_{shift} W(t) \tilde{x}_{365}
\quad\text{and}\quad
F(t) := \log.(G(t))
.$$
Then 
$$
G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x}_{365} = (P_{det}M_{shift} \oslash \tilde{x}_{365}^T) U}_A t
$$
where $\oslash$ is the Kronecker product with reversed arguments, which satisfies $\mathrm{vec}(A B C) = (A \oslash C^T) \mathrm{vec} B$.
Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
$$
\frac{\partial}{\partial t} F(t) =
\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
.$$
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
&= A^T z \\
&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}_{365}) z \\
&= U^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
.
\end{align*}
Since $U$ is unitary, we have $$
\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
UU^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
.$$
The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.

>>>>

\subsection{Fitting the model}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x}_{365})}^2 + \frac{1}{2}\lambda \norm{\log W}^2
.$$
The first term measures the average logarithm of the predicted values $\hat{y}$ (detection events only). The second term measures the average logarithm of the weights.
\begin{proposition}
    Let $\hat{z}$ be the vector $[\hat{z}_1\,\,\dots\,\,\hat{z}_n]^T$ where 
    $$
    \hat{z}_i = \begin{cases}
        \frac{\log \hat{y}_i}{\hat{y}_i} & \text{if observation $i$ is a detection} \\
        0 & \text{if observation $i$ is a non-detection} 
    \end{cases}
    .$$
    Then the gradient of $f(W)$ is the diagonal matrix
    $$
    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left(M_{shift}^T \hat{z} \tilde{x}_{365}^T\right) + \lambda W^{-1} \log W
    $$
    where $\mathrm{Diag}$ is the map that sets the off-diagonal elements of a matrix to zero.
\end{proposition}

\begin{proof}[Proof (still to be checked for calculation errors)]

Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
Define $$
G(t) :=
P_{det} \hat{y}_i
=
P_{det} M_{shift} W(t) \tilde{x}_{365}
\quad\text{and}\quad
F(t) := \log.(G(t))
.$$
Then 
$$
G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x}_{365} = (P_{det}M_{shift} \oslash \tilde{x}_{365}^T) U}_A t
$$
where $\oslash$ is the Kronecker product with reversed arguments, which satisfies $\mathrm{vec}(A B C) = (A \oslash C^T) \mathrm{vec} B$.
Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
$$
\frac{\partial}{\partial t} F(t) =
\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
.$$
Write $z := P_{det} \hat{z}$.
We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(t)$.
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
&= A^T z \\
&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}_{365}) z \\
&= U^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T \\
&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}_{365}^T \\
&= U^T M_{shift}^T z \tilde{x}_{365}^T
\end{align*}
where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
Since $U$ is unitary, we have $$
\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
UU^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
$$
The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.
